Mechanisms of AIAN Inequality in Child Welfare Across U.S. States.

### 2020-12-22 cfp_final_report.Rmd
add discussion

5. Population data

@@

Once a decision has been made to place a Native child into foster care, the Indian Child Welfare Act specifies a preference for placements with kin, with members of the child's tribe, or with another native family. Non-native and non-kin foster families are intended to be a last resort, when preferred placement settings are unavailable. Figure \@ref(fig:non-icwa) shows the proportion of AIAN children in foster care who were ever placed into a non-kin non-native foster setting during their time in foster care. Infants have the highest likelihood of being placed with kin or with a native family. However, if 41 US States, we estimate that more than 50 percent of AIAN infants are placed in a non-native non-kin foster setting. Maryland, New Hampshire, Connecticut, Pennsylvania, North Carolina, and Florida have the highest rate of ICWA preferred placements for infants, ranging between 54 and 61 percent of infants placed with either kin or an AIAN foster family. In Georgia, South Carolina, and Kentucky, more than 80 percent of AIAN infants were placed with a non-kin non-Native foster caretaker. States with the greatest frequency of AIAN foster care also routinely place AIAN children with non-kin non-AIAN caretakers. In South Dakota, 70 percent of infants are placed in non-kin non-Native settings, In Washington and North Dakota, 65 percent of AIAN infants are placed in non-kin non-AIAN settings. In Oklahoma, Minnesota, Wisconsin, and Montana, more than half of all AIAN infants placed in foster care are placed with non-kin non-AIAN caretakers. The likelihood that an AIAN child weill be placed with a family member or other Native foster caretaker typically decreases with age. There is generally a very small chance that an AIAN teenager placed into foster care will be place with kin or with an AIAN foster family.

@@

>>>>

@@

Once a decision has been made to place a Native child into foster care, the Indian Child Welfare Act specifies a preference for placements with kin, with members of the child's tribe, or with another native family. Non-native and non-kin foster families are intended to be a last resort, when preferred placement settings are unavailable. Figure \@ref(fig:non-icwa) shows the proportion of AIAN children in foster care who were ever placed into a non-kin non-native foster setting during their time in foster care. Infants have the highest likelihood of being placed with kin or with a native family. However, if 41 US States, we estimate that more than 50 percent of AIAN infants are placed in a non-native non-kin foster setting. Maryland, New Hampshire, Connecticut, Pennsylvania, North Carolina, and Florida have the highest rate of ICWA preferred placements for infants, ranging between 54 and 61 percent of infants placed with either kin or an AIAN foster family. In Georgia, South Carolina, and Kentucky, more than 80 percent of AIAN infants were placed with a non-kin non-Native foster caretaker. States with the greatest frequency of AIAN foster care also routinely place AIAN children with non-kin non-AIAN caretakers. In South Dakota, 70 percent of infants are placed in non-kin non-Native settings, In Washington and North Dakota, 65 percent of AIAN infants are placed in non-kin non-AIAN settings. In Oklahoma, Minnesota, Wisconsin, and Montana, more than half of all AIAN infants placed in foster care are placed with non-kin non-AIAN caretakers. The likelihood that an AIAN child will be placed with a family member or other Native foster caretaker typically decreases with age. There is generally a very small chance that an AIAN teenager placed into foster care will be place with kin or with an AIAN foster family.

@@

This report has provided two sets of empirical findings.

First, we examine data on the breadth of the separation of AIAN children prior to the passage of the Indian Child Welfare Act, and compare these data to contemporary AIAN family separation through the child welfare system. We find that, while the overall scale of AIAN family separation has decreased since ICWA due to the closing of BIA residential boarding schools, the scale of AIAN family separation through fostering and adoption has increased in the states for which we have comparable data. In some states, like Oklahoma, Washington, California, and Alaska, the frequency of AIAN foster care and adoption has increased dramatically since the mid-1970s.

These results should be interpreted with several limitations in mind, however. Because of changes in the ways that the US Census tabulates AIAN populations, and because of cultural shifts in how AIAN people identify themselves when asked questions about racial identity on surveys like the Census (CITES), estimates of the size of the AIAN child population are not comparable between 1976 and 2014 - 2018. As such, we cannot directly compare population adjusted incidence rates, only raw counts. Nonetheless, the magnitude of growth in the frequency of AIAN foster care and adoption in many states has outpaced the kinds of growth we would expect from demographic processes alone, and suggests that the scope of AIAN fostering and adoption has indeed grown over this period. Future research should explore methods for normalizing population counts across these two periods.

We should also consider AAIA's counts of Native children in foster care and adoption in the pre-ICWA period to be conservative. AAIA relied on agency surveys to count AIAN children that had been separated from their families, and limited survey non-response negatively biases these counts. Additionally, many state and local agencies did not collect data on a child's racial or ethnic identity during this period. In some states, AAIA imputed the total number of AIAN children in foster care and adoption by applying the state or local population composition to the total count of children in foster care or adoption. Assuming any disparity in placement (a reasonable assumption), these imputations would be negatively biased, as AAIA suggests in their report. As such, the pre-ICWA placement numbers are likely conservative estimates of the scale of AIAN family separation in 1976. Because there is likely no such bias in contemporary numbers, estimates of proportional change (as displayed in Figure \@ref(fig:icwa-pct)) may overstate the magnitude of change since ICWA.

Second, we examine data on the contemporary contact of AIAN children with the child welfare system at 5 distinct stages of case processing: investigations, substantiations, removals, placement settings, and termination of parental rights. We find that AIAN children are more likely than white children to be placed into foster care, even when rates of investigation and substantiation of cases are similar to those for white children. We find that inequalities in case outcomes vary dramatically across regions of the country. Inequalities for AIAN children are typically low in the South and Northeast, and typically very high in the Midwest, Great Plains, Mountain West, and Northwest. While some states, such as Alaska, Minnesota, and North and South Dakota exhibit high inequalities in risks of investigation for AIAN children, we find that inequalities in family separation are typically a function of later case processing decisions. In many places, decisions made once a child is in the system, are critical in exacerbating the unequal exposure of AIAN children to family separation.

Efforts to reduce AIAN family separation should be mindful of where inequalities emerge for Native children and families. In some places, the front door of reporting and investigations accounts for a great deal of inequality. In others, AIAN children are treated categorically differently than white children when it comes time to make a decision about removal into foster care after an investigation or substantiation. Our data also show that all states are not adequately addressing ICWA placement preferences. In nearly all states, AIAN children can expect to be placed in a non-kin non-Native setting at some point during their time in foster care.
